Accelerated oblique random survival forests

### 2022-06-20 docs/index.Rmd
adding VI to slides

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 600,
  fig.align='center',
  fig.height=5
)

library(tidyverse)
library(ggforce)
library(rpart)
library(rpart.plot)
library(party)
library(parttree)
library(palmerpenguins)
library(gt)
library(table.glue)

tar_load(
  names = c(
    bm_pred_model_viz,
    bm_pred_time_viz,
    bm_pred_viz,
    data_key
  )
)

@@

    + Decision trees

    + Random survival forests (RSF)

@@

    + Newton Raphson scoring

- Benchmark

    + Datasets & learners

    + Evaluation

    + Results

- Software

    + R package & website

@@

@@

# .larger[Benchmark]

@@

@@

>>>>

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 100,
  fig.align='center',
  fig.height=5
)

library(targets)
library(tidyverse)
library(ggforce)
library(rpart)
library(rpart.plot)
library(party)
library(parttree)
library(palmerpenguins)
library(gt)
library(table.glue)

tar_load(
  names = c(
    bm_pred_model_viz,
    bm_pred_time_viz,
    bm_pred_viz,
    bm_vi_smry,
    data_key
  )
)

@@

@@

- Benchmark of prediction accuracy

- Benchmark of variable importance

- R package & website

@@

## Negation variable importance (VI)

Negation VI is similar to permutation VI in that it measures how much a model’s prediction error increases when a variable’s role in the model is de-stabilized.

- Negation VI measures the increase in an oblique RF's prediction error after flipping the sign of all coefficients linked to a variable.

- As the magnitude of a coefficient increases, so does the probability that negating it will change the oblique RF's predictions.

- Since the coefficients in each non-leaf node of an oblique RFs are adjusted for the accompanying predictors, negation VI may provide better estimation of VI in the presence of correlated variables compared to standard VI techniques.

@@

# .larger[Prediction Benchmark]

@@

# .larger[Variable Importance Benchmark]

## Variable importance techniques

- `randomForestSRC-permutation` - the classic approach proposed by Leo Breiman

- `aorsf-negation` - the new approach I am proposing

- `aorsf-anova` - The only (?) other variable importance technique for oblique random forests

- `aorsf-shap` - SHAP values from an accelerated oblique RSF model

- `xgboost-shap` - SHAP values from an xgboost model (gold standard)

- `xgboost-gain` - Gain importance from an xgboost model (classic)

## Simulated data

We simulated five classes of predictor variables:

- __junk__: no relationship with the outcome.

- __main__: linearly related to the outcome

- __non-linear__: sinusoidal relationship to the outcome

- __combination__: latent variables formed from combinations of the combination variables have a linear relationship to the outcome.

- __interactions__: related to the outcome by multiplicative interaction with one other variable.

## Simulated data

Start with a multivariate normal draw

- All columns have a mean of zero and standard deviation of 1.

- Pairwise correlations are selected at random uniformly between a lower and upper boundary.

- A predictor matrix including interactions, non-linear mappings, and linear combinations is used to generate a time-to-event outcome using the `simsurv` R package

- Transforms are dropped from the matrix after generating the outcome values.

VI techniques are compared based on their discrimination (C-statistic) between relevant and irrelevant variables.

Normally a C-statistic assesses if a predicted score discriminates between positive and negative cases of a binary outcome.

We use the C-statistic differently by treating each variable like a binary outcome (e.g., the outcome is 1 if the variable is relevant, 0 otherwise).

In other words, we assess VI as if it were a ‘prediction’ for the status of the variable. (can only do this for simulated data).

The resulting C-statistic measures the probability that a VI technique will rank a relevant variable higher than an irrelevant variable.

```{r out.width='100%', fig.align='center'}

library(plotly)

fig_data <- bm_vi_smry %>%
  ungroup() %>%
  filter(n_obs == 1000) %>%
  mutate(
    model = recode(
      model, 'randomForestSRC-permutation' = 'rfsrc-permutation'
    ),
    model = fct_reorder(model, .x = overall),
    pred_corr_max = factor(pred_corr_max),
    text = table_glue(
      "{model}<br>\\
          Mean computation time: {as.numeric(time)} seconds<br>\\
          C-statistic by variable class:<br>\\
          Main effect: {main}<br>\\
          Non-linear effect: {nlin}<br>\\
          Combination effect: {cmbn}<br>\\
          Interactions:<br>\\
             w/main effect: {intr_main}<br>\\
             w/non-linear effect: {intr_hidden_nlin}<br>\\
             w/combination effect: {intr_hidden_cmbn}"
    ))

fig <- plot_ly(width = 800, height = 650) %>%
  add_trace(
    type = 'bar',
    x = fig_data$pred_corr_max,
    y = fig_data$overall,
    color = fig_data$model,
    text = table_value(fig_data$overall),
    textposition = 'outside',
    hoverinfo = 'text',
    hovertext = fig_data$text
  ) %>%
  plotly::layout(
    title = list(text = 'Results when N = 1000', x = 0.1),
    xaxis = list(title = 'Max correlation among predictors'),
    yaxis = list(title = 'C-statistic'),
    legend = list(title = list(text = '<b>VI technique</b>'))
  )

fig

```{r out.width='100%', fig.align='center'}

library(plotly)

fig_data <- bm_vi_smry %>%
  ungroup() %>%
  filter(n_obs == 5000) %>%
  mutate(
    model = recode(
      model, 'randomForestSRC-permutation' = 'rfsrc-permutation'
    ),
    model = fct_reorder(model, .x = overall),
    pred_corr_max = factor(pred_corr_max),
    text = table_glue(
      "{model}<br>\\
          Mean computation time: {as.numeric(time)} seconds<br>\\
          C-statistic by variable class:<br>\\
          Main effect: {main}<br>\\
          Non-linear effect: {nlin}<br>\\
          Combination effect: {cmbn}<br>\\
          Interactions:<br>\\
             w/main effect: {intr_main}<br>\\
             w/non-linear effect: {intr_hidden_nlin}<br>\\
             w/combination effect: {intr_hidden_cmbn}"
    ))

fig <- plot_ly(width = 800, height = 650) %>%
  add_trace(
    type = 'bar',
    x = fig_data$pred_corr_max,
    y = fig_data$overall,
    color = fig_data$model,
    text = table_value(fig_data$overall),
    textposition = 'outside',
    hoverinfo = 'text',
    hovertext = fig_data$text
  ) %>%
  plotly::layout(
    title = list(text = 'Results when N = 5000', x = 0.1),
    xaxis = list(title = 'Max correlation among predictors'),
    yaxis = list(title = 'C-statistic'),
    legend = list(title = list(text = '<b>VI technique</b>'))
  )

fig

@@

(https://bcjaeger.github.io/aorsf/)

### 2022-06-20 paper/jmlr/main.Rnw
adding VI to slides

@@

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      dpi = 300)

@@

tar_load(names = c(bm_pred_comb,
                   bm_vi_comb,
                   bm_pred_viz,
                   bm_pred_model_viz,
                   data_key,
                   model_key))

to_omit <- bm_pred_comb |>
  filter(is.na(cstat) | is.na(ibs_scaled)) |>
  distinct(data, run) |>
  as.list() |>
  map(unique)

bm_pred_comb <- bm_pred_comb |>
  filter( ! (data %in% to_omit$data & run %in% to_omit$run) )

@@

\section{Novel techniques for oblique random survival forests}

@@

Negation VI is similar to permutation VI in that it measures how much a model’s prediction error increases when a variable’s role in the model is de-stabilized. More specifically, negation VI measures the increase in an oblique RF's prediction error after flipping the sign of all coefficients linked to a variable (\ie, negating them). As the magnitude of a coefficient increases, so does the probability that negating it will change the oblique RF's predictions. Since the coefficients in each non-leaf node of an oblique RFs are adjusted for the accompanying predictors, negation VI may provide better estimation of VI in the presence of correlated variables compared to standard VI techniques. Although the current article focuses on oblique RSFs, negation VI can be applied to any oblique RF.

@@

\subsubsection{Learners}

@@

@@

Our primary metric for evaluating the accuracy of predicted risk is the integrated and scaled Brier score \citep{graf1999assessment}. Consider a testing data set:
$$\dataset_{\text{test}} = \left\{ (T_i, \delta_i, x_{i}) \right\}_{i=1}^{N_{\text{test}}}.$$
Let $\widehat{S}(t \mid x_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$.
 For observation $i$ in $\dataset_{\text{test}}$, let $\widehat{S}(t \mid \bm{x}_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$. Define \begin{align*}
\bstat = \frac{1}{\ntest} \sum_{i=1}^{\ntest} &\{ \widehat{S}(t \mid \bm{x}_i)^2 \cdot I(T_i \leq t, \delta_i = 1) \cdot \widehat{G}(T_i)^{-1} \\ &+ [1-\widehat{S}(t \mid \bm{x}_i)]^2 \cdot I(T_i > t) \cdot \widehat{G}(t)^{-1}\}
\end{align*} where $\widehat{G}(t)$ is the Kaplan-Meier estimate of the censoring distribution. As $\bstat$ is time dependent, integration over time provides a summary measure of performance over a range of plausible prediction horizons. The integrated $\bstat$ is defined as \begin{equation}
\bsbar = \frac{1}{t_2 - t_1}\int_{t_1}^{t_2} \widehat{\text{BS}}(t) dt.
\end{equation} In our results, $t_1$ and $t_2$ are the 25th and 75th percentile of event times, respectively. $\bsbar$, a sum of squared prediction errors, can be scaled to produce a measure of explained residual variation (\ie, an $R^2$ statistic) by computing \begin{equation}
R^2 = 1 - \frac{\bsbar}{\bskap}
\end{equation} where $\bskap$ is the integrated Brier score when a Kaplan-Meier estimate for survival based on the training data is used as the survival prediction function $\widehat{S}(t)$. We refer to this $R^2$ statistic as the index of prediction accuracy and we scale its values by 100 to avoid unnecessary leading zero's. For example, we present 25 if $R^2$ is 0.25 and present 10.2 if the difference between two $R^2$ is 0.102.

@@

<<fig.height=10, fig.width=9, dpi=400>>=
bm_pred_viz$ibs_scaled
@

<<fig.height=10, fig.width=9, dpi=400>>=
bm_pred_model_viz$ibs_scaled
@

\subsection{Benchmark of variable selection}

@@

<<fig.height=10, fig.width=9, dpi=400>>=
bm_pred_viz$cstat
@

>>>>

\newcommand{\secref}[1]{Section \ref{#1}}

@@

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      dpi = 600)

@@

tar_load(names = c(bm_pred_clean,
                   bm_vi_comb,
                   bm_pred_viz,
                   bm_pred_model_viz,
                   data_key,
                   model_key))

bm_pred_comb <- bm_pred_clean

@@

\section{The acceleracted oblique random survival forest}

@@

Negation VI is similar to permutation VI in that it measures how much a model’s prediction error increases when a variable’s role in the model is de-stabilized. More specifically, negation VI measures the increase in an oblique RF's prediction error after flipping the sign of all coefficients linked to a variable (\ie, negating them). As the magnitude of a coefficient increases, so does the probability that negating it will change the oblique RF's predictions. Since the coefficients in each non-leaf node of an oblique RFs are adjusted for the accompanying predictors, negation VI may provide better estimation of VI in the presence of correlated variables compared to standard VI techniques.

For consistency with prior VI techniques for the RSF and for its computational efficiency, we use Harrell's concordance (C)-statistic \citep{harrell1982evaluating} to measure change in prediction error when computing negation VI. We also note that while the current article focuses on oblique RSFs, negation VI can be applied to any oblique RF and can be applied with any applicable error function.

@@

The aim of this numeric experiment is to evaluate and compare the accelerated oblique RSF with its predecessor (the oblique RSF from the \texttt{obliqueRSF} R package) and with other machine learning algorithms for risk prediction. Inferences drawn from this experiment include equivalence and inferiority tests based on Bayesian hierarchical models.

\subsubsection{Learners} \label{sec:learners}

@@

Our primary metric for evaluating the accuracy of predicted risk is the integrated and scaled Brier score \citep{graf1999assessment}. Consider a testing data set:
$$\dataset_{\text{test}} = \left\{ (T_i, \delta_i, x_{i}) \right\}_{i=1}^{N_{\text{test}}}.$$
Let $\widehat{S}(t \mid x_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$.
 For observation $i$ in $\dataset_{\text{test}}$, let $\widehat{S}(t \mid \bm{x}_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$. Define \begin{align*}
\bstat = \frac{1}{\ntest} \sum_{i=1}^{\ntest} &\{ \widehat{S}(t \mid \bm{x}_i)^2 \cdot I(T_i \leq t, \delta_i = 1) \cdot \widehat{G}(T_i)^{-1} \\ &+ [1-\widehat{S}(t \mid \bm{x}_i)]^2 \cdot I(T_i > t) \cdot \widehat{G}(t)^{-1}\}
\end{align*} where $\widehat{G}(t)$ is the Kaplan-Meier estimate of the censoring distribution. As $\bstat$ is time dependent, integration over time provides a summary measure of performance over a range of plausible prediction horizons. The integrated $\bstat$ is defined as \begin{equation}
\bsbar = \frac{1}{t_2 - t_1}\int_{t_1}^{t_2} \widehat{\text{BS}}(t) dt.
\end{equation} In our results, $t_1$ and $t_2$ are the 25th and 75th percentile of event times, respectively. $\bsbar$, a sum of squared prediction errors, can be scaled to produce a measure of explained residual variation (\ie, an $R^2$ statistic) by computing \begin{equation}
R^2 = 1 - \frac{\bsbar}{\bskap}
\end{equation} where $\bskap$ is the integrated Brier score when a Kaplan-Meier estimate for survival based on the training data is used as the survival prediction function $\widehat{S}(t)$. We refer to this $R^2$ statistic as the index of prediction accuracy \citep{kattan2018index} and we scale its values by 100 to avoid unnecessary leading zero's. For example, we present 25 if $R^2$ is 0.25 and present 10.2 if the difference between two $R^2$ is 0.102.

Our secondary metric for evaluating predicted risk is the time-dependent concordance (C)-statistic. Briefly, we compute the time-dependent C-statistic proposed by \citet{blanche2013estimating}, which is interpreted as the probability that a risk prediction model will assign higher risk to a case (\ie, an observation with $T \leq t$ and $\delta = 1$) versus a non-case (\ie, an observation with $T > t$). Similar to the index of prediction accuracy, observations with $T \leq t$ and $\delta = 0$ only contribute to inverse proportion of censoring weights.

@@

We use a collection of publicly available data sets to benchmark the prediction accuracy and computational efficiency of the accelerated ORSF and each of the other learners described in \secref{sec:learners}

@@

<<fig.height=10, fig.width=9>>=
bm_pred_viz$ibs_scaled
@

<<fig.height=10, fig.width=9>>=
bm_pred_model_viz$ibs_scaled
@

\subsection{Benchmark of variable importance}

The aim of this numeric experiment is to compare negation VI to existing methods for VI. We consider methods that are intrinsic to the oblique RF (\eg ANOVA VI), those that are intrinsic to the RF (\eg permutation VI), and those that are model-agnostic (\eg SHAP VI). All methods are evaluated based on their ability to discriminate between variables that are related to a censored outcome versus those that are not related to it.

\subsubsection{Variable importance techniques}

We compute permutation VI using the \texttt{randomForestSRC} package. Although the \texttt{party} package implements the approach to VI developed by \citet{strobl2007bias}, the developers of the \texttt{party} package note that the implementation of this approach for survival outcomes is ``extremely slow and experimental'' as of version 1.3.10. Therefore, it is not incorporated in the current study's numeric experiments. We compute ANOVA VI using the \texttt{aorsf} package, applying a p-value threshold of 0.01 as recommended by \citet{menze2011oblique}. We compute SHAP VI for boosted tree models using the \texttt{xgboost} package and also compute SHAP VI for accelerated oblique RSFs using the \texttt{fastshap} package.

\subsubsection{Variable types}

We considered five classes of predictor variable, with each class characterized by its variables' relationship to a right-censored outcome. Specifically, \begin{itemize}
\item \textit{junk} variables had no relationship with the outcome.
\item \textit{main effect} variables were linearly related to the outcome
\item \textit{non-linear effect} variables had a sinusoidal relationship to the outcome
\item \textit{combination effect} Latent variables formed by linear combinations of three observed variables. The observed variables have no relationship to the outcome while the latent variable is linearly related to it.
\item \textit{interaction effect} variables related to the outcome by multiplicative interaction with one other variable.
\end{itemize}

\subsubsection{Simulated data}

We generated each set of simulated data with the following steps. First, a $n \times p$ multivariate normal matrix was generated at random, with each of $p$ columns representing one predictor variable and each of $n$ rows one observation. Each predictor variable had a mean of zero, standard deviation of 1, and correlation with other predictor variables drawn at random between a lower and upper boundary. Second, a time-to-event outcome with roughly 45\% of observations censored is generated using the \texttt{simsurv} package. The full predictor matrix (\ie, including interactions, non-linear mappings, and combinations) is used to generate the outcome. Third, interactions, non-linear mappings, and combinations are dropped from the predictor matrix that learning algorithms will use to estimate VI.

\subsubsection{Parameter specifications}

Parameters that varied in the current simulation study included the number of observations (1000, 3000, and 5000) and the minimum and maximum correlation between predictors (-0.1 to 0.1, -0.3 to 0.3, and -0.5 to 0.5). Parameters that remain fixed throughout the study included the number of predictors in each class (15) and the effect size of each predictor, with an increase of one standard deviation associated with a 64\% increase in relative risk.

\subsubsection{Evaluation of variable importance}

We compared VI techniques based on their discrimination (C-statistic) between relevant and irrelevant variables. Normally a C-statistic assesses if a predicted score discriminates between positive and negative cases of a binary outcome. We use the C-statistic in the current simulation study by treating each variable like a binary outcome (e.g., the outcome is 1 if the variable is relevant, 0 otherwise). In other words, we assess VI as if it were a ‘prediction’ for the status of the variable. The resulting C-statistic measures the probability that a VI technique will rank a relevant variable higher than an irrelevant variable.

\section{Discussion}

\subsection{Implications of our results}

\subsection{Limitations and next steps}

@@

<<fig.height=10, fig.width=9>>=
bm_pred_viz$cstat
@
